---
title: "210 Project"
author: "Austin Brown and Ryan Yu"
format: pdf
editor: 
  markdown: 
    wrap: 72
---

```{r, echo = FALSE, message=FALSE, warning=FALSE}
library(tidymodels)
library(tidyverse)
library(Stat2Data)
library(MASS)
library(caret)
library(leaps)
library(glmnet)
library(broom)
library(nnet)
library(ggfortify)
library(naniar)
library(UpSetR)
library(lme4)
library(ggplot2)
options(warn=-1)

```

```{r, echo = FALSE, message=FALSE, warning=FALSE}
player_data <- read.csv(
  "https://query.data.world/s/5zns27g5v3ce74zgwtcosxygtexpef?dws=00000", 
  header=TRUE, stringsAsFactors=FALSE)
player_data <-subset(player_data, select = -c(url,birth_date, X, NCAA_efgpct
            ) )
player_data$career_length <- player_data$active_to - player_data$active_from

breakpoints <- quantile(player_data$NCAA_games, probs = 
                          c(0, 0.25, 0.5, 0.75, 1), na.rm = TRUE)
# Use the cut() function to create the four categories
player_data$NCAA_length <- cut(player_data$NCAA_games, breaks = breakpoints, 
                               labels = c("short", "mid-short", "mid-long", 
                                          "long"))

player_data <- na.omit(player_data)
player_data <- player_data %>% filter(NBA_g_played > 20) %>% 
  filter(NCAA_games > 5) 
```

Position and NCAA Career length Variables

```{r, echo = FALSE, message=FALSE, warning=FALSE}
#test
player_data$position<- substring(player_data$position, first=1, last=1)
```

## Introduction
	For NBA team owners and general managers, the college basketball level 
	represents a key scouting opportunity as they search for their next potential 
	draft prospects. In fact, for the 2020-2021 NBA season, 84.5% of players on 
	NBA rosters played division I college basketball (DNA of Sports). The college 
	level presents players an opportunity to showcase their abilities on both ends
	of the floor, and prove to scouts that they are ready for the next level. 
	However, it’s also important to note that the college game is vastly different 
	from the NBA game in certain ways. In addition to the rules being slightly 
	altered, the NBA game is faster-paced, more spread out, and has a higher 
	concentration of talent on the floor (Brokke). Because of this, a player who 
	fills the stats sheet and dominates at the college level may not be a 
	guaranteed top pick and success in the NBA draft. The goal of this research 
	report is to examine the relationship between college basketball statistics 
	and NBA statistics, specifically with the focus of scoring ability, and 
	ultimately derive an optimal model to predict the NBA’s future prolific 
	scorers based on college performance. 

## Data
The dataset we are using for this project comes from Data World in a table that 
originally contained 34 columns and 4,576 rows. A link to the original data 
dictionary can be found here. This dataset contains observations of people who 
have had both NBA and NCAA basketball careers, with data from the beginning of 
the NBA until 2020. Key identifiers in the dataset include name, date of birth,
and college. For the purpose of this project, however, we are focused on 
performance statistics. Specifically, we are looking at NBA PPG 
(total points/total games played) as a measure of offensive proficiency in the 
NBA. As our main predictors of interest, we are looking at college performance 
statistics. This includes PPG, total games played, and also includes measures of 
efficiency such as field goal percentage (includes both two and three point 
shots), three point field goal percentage, and free throw percentage. All 
shooting percentages are measured as total makes divided by total attempts. For 
the purpose of our research, we eliminated url, birth date, X 
(number identifier) from our dataset. We also removed NCAA effective field goal 
percentage, as this is not a commonly measured statistic at the college level, 
and thus had many missing observations. After we removed these columns, we used 
the na.omit function to remove all observations with NAs in the columns. In 
addition to this, we trimmed the data by filtering for observations with a 
minimum of 6 college games and 21 NBA games, to avoid samples without a 
reasonable sample size. For the purpose of our investigation, we also altered 
the position column to only represent a player’s primary position in the case 
that they were labeled as more than one. For example, if a player was labeled as 
“G-F” (meaning Guard/Forward), we changed their label to just “G”. We also 
created a career length variable by subtracting the player’s starting NBA year 
from their final NBA year. Lastly, we created a new categorical variable, 
“NCAA_length”, to categorize a player’s college career into one of four 
categories: “short”, “mid-short”, “mid-long”, and “long”. We did this by 
splitting the data into quarters given the range of NCAA games in the dataset 
from 16 to 152. 


## EDA

```{r Graph1, echo = FALSE, message=FALSE, warning=FALSE}
player_data %>%
  ggplot(mapping = aes(x = NCAA_ppg, y = NBA_ppg)) +
  geom_point() +
  geom_smooth(method = "lm") + 
  labs(title = "College vs. NBA PPG", 
       x = "NCAA PPG", y = "NBA PPG")
```

Our first initiative was to compare the direct analog to NBA ppg, which
is NCAA ppg. As can be seen in the scatter plot above, there is a direct
positive relationship between NCAA PPG and NBA PPG. One interesting
thing to note, however, is that there are cases of 'high' NBA scoring
averages (> 15) even for players who had NCAA scoring averages of less
than 15 PPG.

```{r Graph2, echo = FALSE, message=FALSE, warning=FALSE}
player_data %>%
  ggplot(mapping = aes(x = NCAA_fgpct, y = NBA_ppg)) +
  geom_point() +
  geom_smooth(method = "lm") + 
  labs(title = "College FG pct vs. NBA PPG", 
       x = "NCAA FG pct", y = "NBA PPG")
```
The second relationship we wanted to explore is between scoring efficiency in 
college (in this case NCAA FG pct) and NBA PPG. We also see a generally positive
relationship between NCAA FG pct and NBA PPG, however, the fitted line in this
plot has a distinctly less extreme slope than in the previous plot. Both of 
these plots suggest a relationship between scoring ability and efficiency in 
college and NBA, however, it is likely that there are more factors involved in
predicting, especially given that at the college level, most players may not
be fully developed from a skill, physical, and mental standpoint. We explore 
this further in the next two graphs. 

```{r Graph3, echo = FALSE, message=FALSE, warning=FALSE}
player_data %>% 
  ggplot(aes(x = NCAA_ppg, y = NBA_ppg, color = NCAA_length)) +
  geom_point() +
  geom_smooth(method = "lm") + 
  facet_wrap(~NCAA_length, nrow = 2, ncol = 4) +
  labs(x = "NCAA PPG", y = "NBA PPG", color = "NCAA Career Length", title = 
         "NCAA PPG vs. NBA PPG, by Length of Time in College")
```

```{r graph4, echo = FALSE, message=FALSE, warning=FALSE}
ggplot(data = player_data, mapping = aes(y = NBA_ppg, x = NCAA_length)) +
  geom_boxplot() +
  labs(x = "NCAA Career Length", y = "NBA PPG", title = 
         "NBA PPG, by Length of Time in College")
```
In the two graphs above, we wanted to get a sense of NBA scoring on a college 
career length basis. What we can see from both plots is that the two shortest
categories for college career (likely players that left early for the NBA) have
observations that spread higher in the NBA ppg category. All four categories 
maintain the upward sloping fitted line in the scatterplot, however, there is 
a clear difference amongst the four categories in the NBA scoring averge 
distributions. This is likely due to the fact that players who have 'lottery 
pick' potential for the NBA draft will often leave college early. These players
are more often the players who develop into the most dangerous scoreres in the 
NBA. As we continue to explore these relationships and begin to craft a model,
college career length will certainly be taken into consideration.

## Methodology
```{r Model Creation, echo = FALSE, message=FALSE, warning=FALSE}
library(glmnet)
set.seed(919)
y <- player_data$NBA_ppg
x <- model.matrix(NBA_ppg ~ NCAA_ppg + NCAA_ft + NCAA__3ptpct + NCAA_fgpct
                + NCAA_length + position*NCAA_ft + 
                    NCAA_length*NCAA_ppg, data = player_data)
m_lasso_cv <- cv.glmnet(x, y, alpha = 1)
best_lambda <- m_lasso_cv$lambda.min
m_best <- glmnet(x, y, alpha = 1, lambda = best_lambda)
m_best$beta
```
### Assumptions Evaluation
```{r, echo = FALSE, message=FALSE, warning=FALSE}
assumptions_model <- lm(NBA_ppg ~ NCAA_ppg + NCAA_ft + NCAA__3ptpct + NCAA_fgpct
                  + NCAA_length + NCAA_length*NCAA_ppg, data = player_data)
assumptions_augment <- augment(assumptions_model)

#residual plot
ggplot(assumptions_augment, aes(x = .fitted, y = .resid)) + 
  geom_point() + 
  geom_hline(yintercept = 0, color = "darkred") + 
  labs(x = "Fitted (predicted) value", y = "Residual") + 
  theme_bw()
```
```{r, echo = FALSE, message=FALSE, warning=FALSE}
#histogram of residuals
ggplot(assumptions_augment, aes(x = .resid)) + 
  geom_histogram(aes(y = ..density..), 
                     fill = "deepskyblue", color = "darkblue") + 
  stat_function(fun = dnorm, 
                args = list(mean = mean(assumptions_augment$.resid),
                            sd = sd(assumptions_augment$.resid)),
                color = "darkred", linewidth = 2) +
  labs(x = "Residual", y = "Density", title = "Histogram of Residuals") + 
  theme_bw()
```
```{r, echo = FALSE, message=FALSE, warning=FALSE}
#qq-plot
ggplot(assumptions_augment, aes(sample = .resid)) +
  stat_qq() + 
  stat_qq_line() + 
  theme_bw() + 
  labs(x = "Theoretical quantiles", 
       y = "Sample quantiles", title = "Q-Q Plot")
```

## Results
```{r Model, echo = FALSE, message=FALSE, warning=FALSE}
NBAppg <- lm(NBA_ppg ~ NCAA_ppg + NCAA_ft + NCAA__3ptpct + NCAA_fgpct
                + NCAA_length + NCAA_length*NCAA_ppg, data = player_data)
tidy(NBAppg)
```
From this model, based off of p-values and a significance level of .05, we can 
identify the following terms as significant: NCAA_ppg, NCAA_3ptpct, 
NCAA_fgpct, and NCAA_lengthmid-long, as all of these terms have p values of less
than 0.05. As for interpretations, if NCAA ppg increases by 1 point, we can 
expect that on average, NBA ppg will increase by 0.398 points while holding all 
other predictors constant. For NCAA_3ptpct, if that increases by 1, or 100%, we 
would expect NBA ppg to increase on average by 1.951 points, while holding all 
other predictors constant. To interpret the percentage terms on a different 
scale, for NCAA_fgpct, for every 1% increase, we would expect on average NBA ppg 
to increase by 0.126 points while holding all other predictors constant. Lastly, 
if a player has a 'mid-long' college career, we would on average expect their 
NBA ppg to be 2.566 lower in comparison to a player with a 'short' college 
career, while holding all other predictors constant.

In terms of answering our research questions, the interpretation of the first 
three terms was logical and in line with our expectations. In general, players 
who score more in college, as well as are more efficient both from the field 
overall but also from three point range are likely to be better scorers in the 
NBA. The last term we interpret is interesting, as it is suggesting that players 
who stay longer in college are expected on average to score less than the 
shortest college career length. This can be explained by the ‘one and done’ 
phenomenon we briefly mentioned earlier, in that the players with the most 
potential NBA upside usually leave college early, even if they haven't fully 
developed yet as a player. In comparison, players who stay in college longer 
may have less natural scoring ability, and thus on average may score less in 
their NBA careers. 

To check the predictive power of our model, we are going to compare our model to
a baseline model, that uses college points per game as the sole predictor as one
of our goals was to evaluate whether NBA General Managers should evaluate player
scoring ability with more than just their college points per game. 

```{r, echo = FALSE, message=FALSE, warning=FALSE}
#model assessment
set.seed(123)
cv_method <- trainControl(method = "cv", number = 10)
baseline <- train(NBA_ppg ~ NCAA_ppg, data = player_data, method = "lm", 
                  trControl = cv_method)
experiment <- train(NBA_ppg ~ NCAA_ppg + NCAA_ft + NCAA__3ptpct + NCAA_fgpct
                + NCAA_length + NCAA_length*NCAA_ppg, 
            data = player_data, method = "lm", trControl = cv_method)
```

### Baseline Model Results: 
```{r, echo = FALSE, message=FALSE, warning=FALSE}
baseline$results[, c("RMSE", "Rsquared", "MAE")]
```
### Experimental Model Results: 
```{r, echo = FALSE, message=FALSE, warning=FALSE}
experiment$results[, c("RMSE", "Rsquared", "MAE")]
```

## Discussion and Conclusion
From the research we did, we can conclude that the most important college 
statistics to consider when predicting NBA scoring average are NCAA ppg, NCAA 3
point percentage, NCAA field goal percentage, and the length of the college 
career for the player. What our research shows is that it is important for 
general managers and NBA scouts to look beyond the basic scoring numbers a 
college player shows. How efficiently a player can score can be an important 
factor, especially because college players transitioning to the NBA may get a 
lower volume of shots at the next level, which would affect their total scoring 
volume. In addition, it is also important to look at the age of a player and how 
long they’ve been in college. Along with consideration of other important 
factors, a young, first year college player with a lot of raw talent may have 
more upside than one of the best college players in the nation that is a 3rd or 
4th year player.  

Predicting NBA points per game from college basketball statistics is a 
challenging task due to several potential limitations. First and foremost, it is 
notoriously difficult to predict individual player performance with a high 
degree of accuracy. If it were easy, there would be no such thing as "draft 
busts," where players selected with high expectations fail to live up to their 
potential. Moreover, there are numerous external factors that can significantly 
affect a player's career trajectory. For instance, injuries, coaching changes,
changes in team roles, and off-court distractions can all impact a player's 
ability to perform at their best. Additionally, the transition from college 
basketball to the NBA is itself a challenging process that can take time to 
adjust to, which can further complicate efforts to predict future performance 
based on college statistics alone. Therefore, while college basketball 
statistics can be useful in predicting NBA success to some extent, it is 
essential to recognize the limitations of this approach and to consider other 
factors that may impact player performance.

In terms of ideas for future work, we can look to incorporate other types of 
statistics such as defensive metrics, as well as more general basketball metrics 
such as on/off ratings. This would allow us to assess college players from a 
more holistic perspective, as scoring is obviously only one part of a total 
player’s success and impact on a team. In addition, we could look to categorize
the type of school that players come from, and investigate the impact of that 
on their future NBA success as well as on how reliable college statistics are 
for predicting NBA success. High-major players from schools like Duke and 
Kentucky, for example, may be easier to predict than players from less-known, 
smaller schools.

## Sources

“What Percent of NBA Players Played in College?” DNA of SPORTS, 24 Feb. 2022, https://www.dnaofsports.com/basketball/what-percent-of-NBA-players-played-in-
college/. 

Brokke, Nathan. “NCAA Mania: Why NBA Is Much Better Basketball than College 
Hoops.” Bleacher Report, Bleacher Report, 3 Oct. 2017, https://bleacherreport.com/articles/636365-ncaa-mania-why-nba-is-much-better-
basketball-than-college-hoops. 

“NBA NCAA Comparisons - Project by BGP12.” Data.world, 6 Feb. 2020, https://data.world/bgp12/nbancaacomparisons/workspace/data-dictionary. 

